---
title: 'Assignment: Local (alpha) Diversity'
author: 'Ashwini Ramesh; Z620: Quantitative Biodiversity, Indiana University'
output: html_document
geometry: margin=2.54cm
---

## OVERVIEW

In this exercise, we will explore aspects of local or site-specific diversity, also known as alpha ($\alpha$) diversity. 
First we will quantify two of the fundamental components of ($\alpha$) diversity: **richness** and **evenness**.
From there, we will then discuss ways to integrate richness and evenness, which will include univariate metrics of diversity along with an investigation of the **species abundance distribution (SAD)**.

## Directions:
1. Change "Student Name" on line 3 (above) with your name.
2. Complete as much of the exercise as possible during class; what you do not complete in class will need to be done on your own outside of class.
3. Use the handout as a guide; it contains a more complete description of data sets along with the proper scripting needed to carry out the exercise.
4. Be sure to **answer the questions** in this exercise document; they also correspond to the handout.
Space for your answer is provided in this document and indicated by the ">" character.
If you need a second paragraph be sure to start the first line with ">".
5. Before you leave the classroom, **push** this file to your GitHub repo.
6. For homework, follow the directions at the bottom of this file. 
7. When you are done, **Knit** the text and code into a PDF file.
8. After Knitting, please submit the completed exercise by creating a **pull request** via GitHub.
Your pull request should include this file `alpha_assignment.Rmd` and the PDF output of `Knitr` (`alpha_assignment.pdf`).


## 1) R SETUP

In the R code chunk below, please provide the code to: 
1) Clear your R environment,
2) Print your current working directory,
3) Set your working directory to your `/Week2-Alpha` folder, and
4) Load the `vegan` R package (be sure to install if needed).

```{r}
rm(list=ls())
getwd()
setwd("C:/Users/DELL/GitHub/QB2017_Ramesh/Week2-Alpha")
#install.packages("vegan")
require("vegan")
```

## 2) LOADING DATA

In the R code chunk below, do the following:
1) Load your dataset, and 
2) Display the structure of the dataset (if the structure is long, use `max.level=0` to show just basic information).

```{r}
data(BCI)
str(BCI, max.level =0)
```

## 3) SPECIES RICHNESS

**Species richness (S)** is simply the number of species in a system or the number of species observed in a sample.

### Observed Richness

In the R code chunk below, do the following:

1. Write a function called `S.obs` to calculate observed richness

2. Use your function to determine the number of species in `site1`, and

3. Compare the output of your function to the output of the `specnumber()` function in vegan.

```{r}
#1. Write a function called `S.obs` to calculate observed richness
S.obs <- function(x = ""){rowSums(x >0) * 1 }

#2. Use your function to determine the number of species in `site1`, and
S.obs(BCI)

#3. Compare the output of your function to the output of the `specnumber()` function in vegan.
specnumber(BCI) #Both funtion have the same value
```

***Question 1***: Does `specnumber()` from `vegan` return the same value for observed richness in `site1` as our function `S.obs`?
What is the species richness of the first 4 sites (i.e., rows) of the BCI matrix?

> ***Answer 1***:Yes, it does.
The species richness is Site 1: 93, Site 2:84, Site 3: 90, Site 4: 94

### Coverage. How Well Did You Sample Your Site?

In the R code chunk below, do the following:

1. Write a function to calculate Good's Coverage, and

2. Use that function to calculate coverage for all sites in the BCI matrix.

```{r}
#1. Write a function to calculate Good's Coverage, and
C <- function(x=""){1 - (rowSums(x==1)/ rowSums(x))}
C(BCI)

#2. Use that function to calculate coverage for all sites in the BCI matrix.
singleton <- function(x = ""){rowSums(x == 1)}
singleton(BCI[1,])/length(BCI[1,])

```
  
***Question 2***: Answer the following questions about coverage:

a.  What is the range of values that can be generated by Good's Coverage? 
b.  What would we conclude from Good's Coverage if $n_{i}$ equaled *N*?
c.  What portion of taxa in `site1` were represented by singletons? 
d.  Make some observations about coverage at the BCI plots.

> ***Answer 2a***:  0 to 1

> ***Answer 2b***:  C = 0, which would mean that there is poor coverage across site

> ***Answer 2c***:  0.13778 ((C[1]/S.obs[1])-1)*(-1) or sum(BCI[1,])/S.obs[1]

> ***Answer 2d***:  It appears that almost all the plots have good coverage (>0.90)

### Estimated Richness

In the R code chunk below, do the following:

1. Load the microbial dataset (located in the `/Week2-Alpha/data` folder),

2. Transform and transpose the data as needed (see handout),

3. Create a vector (`soilbac1`) with the bacterial OTU abundances at any site in the dataset,

4. Calculate the observed richness at that particular site, and 

5. Calculate the coverage at that particular site

```{r}
soilbac <- read.table("data/soilbac.txt", sep ="\t", header = TRUE, row.names =1)
str(soilbac)
soilbac.t <- as.data.frame(t(soilbac))
soilbac1 <- soilbac.t[1,]
S.obs(soilbac1) #1074
C(soilbac1) #0.6479
```

***Question 3***: Answer the following questions about the soil bacterial dataset. 

a.  How many sequences did we recover from the sample `soilbac1`, i.e. *N*? 
b.  What is the observed richness of `soilbac1`? 
c.  How does coverage compare between the BCI sample (`site1`) and the KBS sample (`soilbac1`)? 

> ***Answer 3a***:  13310

> ***Answer 3b***: 1074

> ***Answer 3c***:  BCI coverage is 0.93, while in soilbac is 0.64. There is 7% more species in BCI than soilbac


### Richness Estimators

In the R code chunk below, do the following:

1. Write a function to calculate **Chao1**,

2. Write a function to calculate **Chao2**, 

3. Write a function to calculate **ACE**, and

4. Use these functions to estimate richness at both `site1` and `soilbac1`. 

```{r}
S.chao1 <- function(x=""){S.obs(x) + sum(x==1)^2/(2*sum(x==2))}

S.chao2 <- function(site = "", SbyS = ""){
SbyS = as.data.frame(SbyS)
x = SbyS[site, ]
SbyS.pa <- (SbyS > 0) * 1
Q1 = sum(colSums(SbyS.pa) == 1)
Q2 = sum(colSums(SbyS.pa) == 2) 
S.chao2 = S.obs(x) + (Q1^2)/(2 * Q2) 
return(S.chao2)
}

S.ace <- function(x = "", thresh = ""){
x <- x[x>0] # excludes zero-abundance taxa
S.abund <- length(which(x > thresh)) # richness of abundant taxa
S.rare <- length(which(x <= thresh)) # richness of rare taxa
singlt <- length(which(x == 1)) # number of singleton taxa
N.rare <- sum(x[which(x <= thresh)]) # abundance of rare individuals
C.ace <- 1 - (singlt / N.rare) # coverage (prop non-singlt rare inds)
i <- c(1:thresh) # threshold abundance range
count <- function(i, y){ # counter to go through i range
length(y[y == i])
}
a.1 <- sapply(i, count, x) # number of individuals in richness i richness classes
f.1 <- (i * (i - 1)) * a.1 # k(k-1)kf sensu Gotelli
G.ace <- (S.rare/C.ace)*(sum(f.1)/(N.rare*(N.rare-1)))
S.ace <- S.abund + (S.rare/C.ace) + (singlt/C.ace) * max(G.ace,0)
return(S.ace)
}

#Chao1 for BCI and soilmicro
S.chao1bci <- S.chao1(BCI[1,])
S.chao1soilbac <- S.chao1(soilbac1)

#Chao2 for BCI and soilmicro
S.chao2bci <- S.chao2(1, BCI) #104.6053
S.chao2soilbac <- S.chao2("T1_1", soilbac.t) #21055.39


#ACE for BCI and soilmicro
S.aceBCI <- S.ace(BCI[1,],10) #4465.983
S.acesoilbac <- S.ace(soilbac1,10) #159.3404

```

### Rarefaction

In the R code chunk below, please do the following:

1. Calculate observed richness for all samples in `soilbac`,

2. Determine the size of the smallest sample,

3. Use the `rarefy()` function to rarefy each sample to this level,

4. Plot the rarefaction results, and

5. Add the 1:1 line and label.


```{r}
#1. Calculate observed richness for all samples in `soilbac`,
soilbac.t <- as.data.frame(t(soilbac))
soilbac.S <- S.obs(soilbac.t)

#2. Determine the size of the smallest sample
min.N <- min(rowSums(soilbac.t)) #Answer: 2119

#3. Use the `rarefy()` function to rarefy each sample to this level
S.rarefy <- rarefy(x = soilbac.t, sample = min.N, se = TRUE)

#4. Plot the rarefaction results
rarecurve(x = soilbac.t, step = 20, col = "blue", cex = 0.6, las=1)
abline(0, 1, col = 'red')
text(1500, 1500, "1:1", pos = 2, col = 'red')
```

***Question 4***: What is the difference between ACE and the Chao estimators?

> ***Answer 4***: The rare species in the Chao estimators are inferred via singletons and doubletons, however in ACE estimators one can set the minimum threshold below which one classififes all the species as rare. These are not restricted to singletons or doubletons as in Chao.

##4) SPECIES EVENNESS
Here, we consider how abundance varies among species, that is, **species evenness**.

### Visualizing Evenness: The Rank Abundance Curve (RAC)
One of the most common ways to visualize evenness is in a **rank-abundance curve** (sometime referred to as a rank-abundance distribution or Whittaker plot).
An RAC can be constructed by ranking species from the most abundant to the least abundant without respect to species labels (and hence no worries about 'ties' in abundance). 

In the R code chunk below, do the following:

1. Write a function to construct a RAC,

2. Be sure your function removes species that have zero abundances, 

3. Order the vector (RAC) from greatest (most abundant) to least (least abundant), and 

4. Return the ranked vector

```{r}

#1. Write a function to construct a RAC
RAC <- function(x=""){
x = as.vector(x)
x.ab = x[x>0] #Be sure your function removes species that have zero abundances
x.ab.ranked = x.ab[order(x.ab, decreasing = TRUE)]
return(x.ab.ranked)
}

```

Now, let's examine the RAC for `site1` of the BCI data set.

In the R code chunk below, do the following:

1. Create a sequence of ranks and plot the RAC with natural-log-transformed abundances,

2. Label the x-axis "Rank in abundance" and the y-axis "log(abundance)"

```{r}
plot.new()
site1 <- BCI[1, ]
rac <- RAC(x = site1)
ranks <- as.vector(seq(1, length(rac)))
opar <- par(no.readonly = TRUE) # Saves default plot parameters
par(mar = c(5.1, 5.1, 4.1, 2.1)) # New settings for par
plot(ranks, log(rac), type = 'p', axes = F, # Plots w/o axes
xlab = "Rank in abundance", ylab = "log(Abundance)",
las = 1, cex.lab = 1.4, cex.axis = 1.25)
box() # Manually adds border
axis(side = 1, labels = T, cex.axis = 1.25) # Manually adds X-axis
axis(side = 2, las = 1, cex.axis = 1.25, # Manually adds Log-Scaled Y-axis
labels = c(1, 2, 5, 10, 20), at = log(c(1, 2, 5, 10, 20)))

```

***Question 5***: What effect does visualizing species abundance data on a log-scaled axis have on how we interpret evenness in the RAC?

> ***Answer 5***: The log scale allows us to easily visualize data points with large differences between them in the same plot. For instance, one can have many 0's and 1's and say, values as high as 100000 in the same data set. Plotting, visualizign and interpreting these raw values can be cumbersome and observing patterns can be pretty tedious. The ln-scale converts this into a simple range between 0 and 11, allowing us to visiualize the same. 


Now that we have visualized unevennes, it is time to quantify it using Simpson's evenness ($E_{1/D}$) and Smith and Wilson's evenness index ($E_{var}$).

### Simpson's evenness ($E_{1/D}$)

In the R code chunk below, do the following:

1. Write the function to calculate $E_{1/D}$, and

2. Calculate $E_{1/D}$ for `site1`.

```{r}
SimpE <- function(x=""){
S <- S.obs(x)
x = as.data.frame(x)
D <- diversity(x,"inv")
E <- (D)/S
return(E)
}

site1 <- BCI[1,]
SimpE(site1)

```

### Smith and Wilson's evenness index ($E_{var}$)

In the R code chunk below, please do the following:

1. Write the function to calculate $E_{var}$,

2. Calculate $E_{var}$ for `site1`, and

3. Compare $E_{1/D}$ and $E_{var}$.

```{r}
#1. Write the function to calculate $E_{var}$,
Evar <- function(x){
x <- as.vector(x[x>0])
1 - (2/pi) * atan(var(log(x)))
}

#2. Calculate $E_{var}$ for `site1`, and
Evar(site1)

#3. Compare $E_{1/D}$ and $E_{var}$.
SimpE(site1)
Evar(site1)

```

***Question 6***: Compare estimates of evenness for `site1` of BCI using $E_{1/D}$ and $E_{var}$.
Do they agree? If so, why? If not, why? What can you infer from the results.

> ***Answer 6***:No, the two indices do not yield the same value. Simpson's index is biased to the most abundance species, while since Evar is log transformed it decreases the bias towards the most abundant species. Since there is a descrepancy in the two values, we can infer that Evar is more sensitive to the large abundance in species and better depicts the evenness in the community.


##5) INTEGRATING RICHNESS AND EVENNESS: DIVERSITY METRICS

So far, we have introduced two primary aspects of diversity, i.e., richness and evenness.
Here, we will use popular indices to estimate diversity, which explicitly incorporate richness and evenness
We will write our own diversity functions and compare them against the functions in `vegan`. 

### Shannon's diversity (a.k.a., Shannon's entropy)

In the R code chunk below, please do the following:

1. Provide the code for calculating H' (Shannon's diversity),

2. Compare this estimate with the output of `vegan`'s diversity function using method = "shannon".

```{r}

#1. Provide the code for calculating H' (Shannon's diversity)
ShanH <- function(x = ""){
H = 0
for (n_i in x){
if(n_i > 0) {
p = n_i / sum(x)
H = H - p*log(p)
}
}
return(H)
}

diversity(site1, index = "shannon")
ShanH(site1)
#They are the same
```

### Simpson's diversity (or dominance)

In the R code chunk below, please do the following:

1. Provide the code for calculating D (Simpson's diversity),

2. Calculate both the inverse (1/D) and 1 - D,

2. Compare this estimate with the output of `vegan's` diversity function using method = "simp".

```{r}
SimpD <- function(x = ""){
D = 0
N = sum(x)
for (n_i in x){
D = D + (n_i^2)/(N^2)
}
return(D)
}

D.inv <- 1/SimpD(site1)
D.sub <- 1-SimpD(site1)

diversity(site1, index = "simp")
D.sub
```

***Question 7***: Compare estimates of evenness for `site1` of BCI using $E_{H'}$ and $E_{var}$.
Do they agree? If so, why? If not, why? What can you infer from the results.
```{r}
ShanH(site1) #Shannon's Index (H')
Evar(site1) #Evar Index

#No, the two indices do not agree. Shannon's Diversity index integrates both diversity and evenness in the community, while Evar calculates eveness alone. One cannot directly infer the evenness of a community from this measure. Instead, one can use Pielou's evenness index(J), which is H'/H'max, where H'max is the maximum possible value of H' (or as ln(total number of species))

Hmaxsite1 <- log(length(site1))
J <- ShanH(site1)/Hmaxsite1

Evar(site1)#0.4735552
J #0.7105613

#These results indicate that Evar and J also do not agree. This is perhaphs because Evar using log of abundances and hences decreases bias at extreme abundances.

```

> ***Answer 7***:

### Fisher's $\boldsymbol\alpha$

In the R code chunk below, please do the following:

1. Provide the code for calculating Fisher's $\boldsymbol\alpha$,

2. Calculate Fisher's $\boldsymbol\alpha$ for `site1` of BCI.

```{r}
rac <- as.vector(site1[site1 > 0])
invD <- diversity(rac, "inv")
invD
# 39.41555
Fisher <- fisher.alpha(rac)
Fisher
# 35.67297

```

***Question 8***: How is Fisher's $\boldsymbol\alpha$ different from $E_{H'}$ and $E_{var}$? What does Fisher's $\boldsymbol\alpha$ take into account that $E_{H'}$ and $E_{var}$ do not?

> ***Answer 8***:While Evar is used to estimate evenness in a community and Shannon's H' integrates both richnes and evenness, and both are are used for calculting a metric of diversity, Fisher's Alpha is used for estimating diversity. The latter takes into account the sampling error in a community, while the former do not. 

##6) MOVING BEYOND UNIVARIATE METRICS OF $\alpha$ DIVERSITY

The diversity metrics that we just learned about attempt to integrate richness and evenness into a single, univariate metric.
Although useful, information is invariably lost in this process. 
If we go back to the rank-abundance curve, we can retrieve additional information -- and in some cases -- make inferences about the processes influencing the structure of an ecological system.

## Species abundance models

The RAC is a simple data structure that is both a vector of abundances.
It is also a row in the site-by-species matrix (minus the zeros, i.e., absences). 

Predicting the form of the RAC is the first test that any biodiversity theory must pass and there are no less than 20 models that have attempted to explain the uneven form of the RAC across ecological systems.

In the R code chunk below, please do the following:

1. Use the `radfit()` function in the `vegan` package to fit the predictions of various species abundance models to the RAC of `site1` in BCI,

2. Display the results of the `radfit()` function, and 

3. Plot the results of the `radfit()` function using the code provided in the handout.

```{r}
#1. Use the `radfit()` function in the `vegan` package to fit the predictions of various species abundance models to the RAC of `site1` in BCI,
RACresults <- radfit(site1)

#2. Display the results of the `radfit()` function, and 
RACresults

#Plot the results of the `radfit()` function using the code provided in the handout
plot.new()
plot(RACresults, las = 1, cex.lab = 1.4, cex.axis = 1.25)
```


***Question 9***: Answer the following questions about the rank abundance curves:
a) Based on the output of `radfit()` and plotting above, discuss which model best fits our rank-abundance curve for `site1`?
b) Can we make any inferences about the forces, processes, and/or mechanisms influencing the structure of our system, e.g., an ecological community?

> ***Answer 9a***:  Visually, one can see that the Mandelbrot model best fits the data points. This is supported by the results of the radfit function that shows that Mandelbrot has the lowest AIC value among all the models.
> ***Answer 9b***:It does not seem feasible to directly make any inferences about the different proccesses influencing the ecological community.


***Question 10***: Answer the following questions about the preemption model:
a.  What does the preemption model assume about the relationship between total abundance (*N*) and total resources that can be preempted?
b.  Why does the niche preemption model look like a straight line in the RAD plot?

> ***Answer 10a***:  The model assumes that resources that can be prempted is a decaying funtion reprsented with (1 - Alpha)^(r-1), and continues to decay until the total abundance (N) is 0
> ***Answer 10b***:The niche preemtion model looks like a straight line because the natural log of abundance is plotted here and natural log(ar) = constant x log alpha(1-alpha) --> which would be a linear function


***Question 11***: Why is it important to account for the number of parameters a model uses when judging how well it explains a given set of data?

> ***Answer 11***: The more number of paramteres one has in the model, there is greater accountability of all the different variables that can correctly predict your data. However, overfitting can become a problem - where one fits in too many variables, wherein there is danger of fitting to nois instead of the actual signal. In fact, Burnham and Anderson estmate tath number of parameters that can estimated should be < m/15; Crawley recommends < m/5; where m is the total sample size.


## SYNTHESIS

1.  As stated by Magurran (2004) the  ${D = } \sum p_i^2$ derivation of Simpson's Diversity only applies to communities of infinite size. For anything but an infinitely large community, Simpson's Diversity index is calculated as ${D = } \sum \frac{n_i(n_i -1)} {N(N-1)}$. 
Assuming a finite community, calculate Simpson's D, 1 - D, and Simpson's inverse (i.e. 1/D) for `site 1` of the BCI site-by-species matrix. 

```{r}
SimpD <- function(x = ""){
D = 0
N = sum(x)
for (n_i in x){
D = D + (n_i^2)/(N^2)
}
return(D)
}

SimpD(site1) #Simpsons D
D.sub <- 1-SimpD(site1) #1 - D
D.inv <- 1/SimpD(site1) #Simpson's inverse
```


2.  Along with the rank-abundance curve (RAC), another way to visualize the distribution of abundance among species is with a histogram (a.k.a., frequency distribution) that shows the frequency of different abundance classes.
For example, in a given sample, there may be 10 species represented by a single individual, 8 species with two individuals, 4 species with three individuals, and so on.
In fact, the rank-abundance curve and the frequency distribution are the two most common ways to visualize the species-abundance distribution (SAD) and to test species abundance models and biodiversity theories.
To address this homework question, use the R function **hist()** to plot the frequency distribution for `site 1` of the BCI site-by-species matrix, and describe the general pattern you see.

```{r}
aa <- hist(as.numeric(BCI[1,]),xlab = "Abundance", ylim = c(0,200), xlim = c(0,28),main = "Frequencey Distribution of Species abundances")
aa$counts #Gives the counts of the freq. disttibution plotted

#Overall, the frequencey distribution seems to peak for abundances < 3, falls steeply and then tapers off. There appears to be a large number of singletons and doubletons in the population (Order of about ~180). Species higher in abundance are few in number (< 15) 
```

3.  We asked you to find a biodiversity dataset with your partner.
This data could be one of your own or it could be something that you obtained from the literature. 
Load that dataset.
How many sites are there?
How many species are there in the entire site-by-species matrix?
Any other interesting observations based on what you learned this week?
> ***Answer 11***: 
dat <- read.csv("murray.csv")
str(dat)
#How many sites are there? 397
#How many species are there in the entire site-by-species matrix? 154
#Any other interesting observations based on what you learned this week? I believe we need to do more detailed analysis to comment in this


## SUBMITTING YOUR ASSIGNMENT
Use Knitr to create a PDF of your completed alpha_assignment.Rmd document, push it to GitHub, and create a pull request.
Please make sure your updated repo include both the HTML and RMarkdown files.

Unless otherwise noted, this assignment is due on **Wednesday, January 25^th^, 2015 at 12:00 PM (noon)**.