---
title: "Assignment: Spatial Diversity"
author: "Ashwini Ramesh; Z620: Quantitative Biodiversity, Indiana University"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW
This assignment will emphasize primary concepts and patterns associated with spatial diversity, while using R as a Geographic Information Systems (GIS) environment. 
Complete the assignment by refering to examples in the handout.

After completing this assignment you will be able to:  
1.  Begin using R as a geographical information systems (GIS) environment.  
2.  Identify primary concepts and patterns of spatial diversity.  
3.  Examine effects of geographic distance on community similarity.  
4.  Generate simulated spatial data.

## Directions:
1. Change "Student Name" on line 3 (above) with your name.
2. Complete as much of the assignment as possible during class; what you do not complete in class will need to be done on your own outside of class.
3. Use the handout as a guide; it contains a more complete description of data sets along with the proper scripting needed to carry out the assignment.
4. Be sure to **answer the questions** in this assignment document.
Space for your answer is provided in this document and indicated by the ">" character.
If you need a second paragraph be sure to start the first line with ">".
5. Before you leave the classroom, **push** this file to your GitHub repo.
6. When you are done wit the assignment, **Knit** the text and code into an html file.
7. After Knitting, please submit the completed assignment by creating a **pull request** via GitHub.
Your pull request should include this file *spatial_assignment.Rmd* and the html output of `Knitr` (*spatial_assignment.html*).


## 1) R SETUP
In the R code chunk below, provide the code to:  

1. Clear your R environment
2. Print your current working directory,
3. Set your working directory to your "*/Week4-Spatial*" folder, and

```{r, results = 'hide'}
#rm(list=ls())
#getwd() 
#setwd("~/GitHub/QuantitativeBiodiversity/QB-2017/Week4-Spatial")

```

## 2) LOADING R PACKAGES

In the R code chunk below, do the following:  

1. Install and/or load the following packages: `vegan`, `sp`, `gstat`, `raster`, `RgoogleMaps`, `maptools`, `rgdal`, `simba`, `gplots`, `rgeos`

```{r, results = 'hide', message = FALSE, warning = FALSE} 
require(vegan)
install.packages('sp') # Classes and methods for handling spatial data require(sp)
require(sp)
install.packages('ggmap')
require(ggmap)
install.packages('gstat') # Methods for geostatistical analyses
require(gstat)
install.packages('raster') # Methods to create a RasterLayer object require(raster)
install.packages('RgoogleMaps') # For querying the Google server for static maps. require(RgoogleMaps)
install.packages('maptools') # Tools for manipulating and reading geospatial data require(maptools)
install.packages('rgdal') # Geospatial Data Abstraction Library
require(rgdal)
install.packages('simba') # Similarity measures for community data
require(simba)
install.packages('gplots') # Programming tools for plotting data
require(gplots)
install.packages('rgeos') # Geostatistical package, used here for semivariograms require(rgeos)
require(rgdal)


```

***Question 1***:  What are the packages `simba`, `sp`, and `rgdal` used for?

> ***Answer 1***:  simba is used to measure similarity measures for community data. sp for the classes and methods for handling spatial data and rgdal for geosatial data absraction library


## 3) LOADING DATA
In the R code chunk below, use the example in the handout to do the following:  

1. Load the Site-by-Species matrix for the Indiana ponds datasets: BrownCoData/SiteBySpecies.csv
2. Load the Environmental data matrix: BrownCoData/20130801_PondDataMod.csv
3. Assign the operational taxonomic units (OTUs) to a variable 'otu.names'
4. Remove the first column (i.e., site names) from the OTU matrix.

```{r, results = 'hide', message = FALSE, warning = FALSE} 
Ponds <- read.table(file = "BrownCoData/20130801_PondDataMod.csv", head = TRUE, sep = ",") 
OTUs <- read.csv(file = "BrownCoData/SiteBySpecies.csv", head = TRUE, sep = ",")
otu.names <- names(OTUs) # Get the names of the OTUs
OTUs <- as.data.frame(OTUs[-1]) # remove first column (site names)

```

***Question 2a***:  How many sites and OTUs are in the SiteBySpecies matrix? 

> ***Answer 2a***:  Sites = 51, OTUs = 16383

***Question 2b***:  What is the greatest species richness found among sites?

> ***Answer 2b***:  


## 4) GENERATE MAPS

In the R code chunk below, do the following:  

1. Using the example in the handout, visualize the spatial distribution of our samples with a basic map in RStudio using the `GetMap` function in the package `RgoogleMaps`. 
This map will be centered on Brown County, Indiana (39.1 latitude, -86.3 longitude).

```{r}
lats <- as.numeric(Ponds[, 3]) # latitudes (north and south) 
lons <- as.numeric(Ponds[, 4]) # longitudes (east and west)
newmap <- GetMap(center = c(39.1,-86.3), zoom = 10, destfile = "PondsMap.png", maptype="terrain")
PlotOnStaticMap(newmap, zoom = 10, cex = 2, col = 'blue') # Plot map in RStudio PlotOnStaticMap(newmap, lats, lons, cex = 1, pch = 20, col = 'red', add = TRUE)
PlotOnStaticMap(newmap, lats, lons, cex = 1, pch = 20, col = 'red', add = TRUE)
```

***Question 3***:  Briefly describe the geographical layout of our sites.

> ***Answer 3***:  


In the R code chunk below, do the following:  

1. Using the example in the handout, build a map by combining lat-long data from our ponds with land cover data and data on the locations and shapes of surrounding water bodies.

```{r}
library(raster)
library(RgoogleMaps)
require(vegan)
#install.packages('sp') # Classes and methods for handling spatial data 
require(sp)
#install.packages('gstat') # Methods for geostatistical analyses
require(gstat)
#install.packages('raster') # Methods to create a RasterLayer object 
require(raster)
#install.packages('RgoogleMaps') # For querying the Google server for static maps. 
require(RgoogleMaps)
#install.packages('maptools') # Tools for manipulating and reading geospatial data 
require(maptools)
#install.packages('rgdal') # Geospatial Data Abstraction Library
require(rgdal)
#install.packages('simba') # Similarity measures for community data
require(simba)
#install.packages('gplots') # Programming tools for plotting data
require(gplots)
#install.packages('rgeos') # Geostatistical package, used here for semivariograms 
require(rgeos)
require(rgdal)

# 1. Import TreeCover.tif as a raster file.
Tree.Cover <- raster("TreeCover/TreeCover.tif") # import TreeCover.tif as a raster file.
# 2. Plot the % tree cover data
plot(Tree.Cover, xlab = 'Longitude', ylab = 'Latitude', main = 'Map of geospatial data for % tree cover, \nwaterbodies, and sample sites')
# 3. Import water bodies as a shapefile.
Water.Bodies <- readShapeSpatial("water/water.shp")
# 4. Plot the water bodies around our study area, i.e., Monroe County.
plot(Water.Bodies, border = 'cyan', axes = TRUE, add = TRUE)
# 5. Convert lat-long data for ponds to georeferenced points.
Refuge.Ponds <- SpatialPoints(cbind(lons,lats))
# 6. Plot the refuge pond locations
plot(Refuge.Ponds, line = 'r', col = "red", pch = 20, cex = 1.5, add = TRUE)
```

***Question 4a***: What are datums and projections?

> ***Answer 4a***:  


## 5) UNDERSTANDING SPATIAL AUTOCORRELATION

***Question 5***: In your own words, explain the concept of spatial autocorrelation.

> ***Answer 5***:


## 6) EXAMINING DISTANCE-DECAY

***Question 6***: In your own words, explain what a distance decay pattern is and what it reveals.

> ***Answer 6***:

In the R code chunk below, do the following:  

1. Generate the distance decay relationship for bacterial communities of our refuge ponds and for some of the environmental variables that were measured. 
Note: You will need to use some of the data transformations within the *semivariogram* section of the handout.

```{r, results = 'hide', message = FALSE, warning = FALSE}
# Construct a new dataframe for coordinates
xy <- data.frame(env = Ponds$TDS, pond.name = Ponds$Sample_ID, lats = Ponds$lat, lons = Ponds$long)
coordinates(xy) <- ~lats+lons # Transform 'xy' into a spatial points dataframe
# Identify the current projection (i.e., lat-long) and datum (NAD83). In our case, the projection 
proj4string(xy) <- CRS("+proj=longlat +datum=NAD83") # coordinate reference system (CRS)
# Then, transform the projection and data so we can get meaningful georeferenced distances 
UTM <- spTransform(xy, CRS("+proj=utm +zone=51 +ellps=WGS84")) # coordinate reference syst
UTM <- as.data.frame(UTM)
xy$lats_utm <- UTM[,2]
xy$lons_utm <- UTM[,3]
env.vgm <- variogram(env~1, data = xy)
plot(env.vgm)






# 1) Calculate Bray-Curtis similarity between plots using the `vegdist()` function
comm.dist <- 1 - vegdist(OTUs)

# 2) Assign UTM lattitude and longitude data to 'lats' and 'lons' variables
lats <- as.numeric(xy$lats_utm)
lons <- as.numeric(xy$lons_utm)

# 3) Calculate geographic distance between plots and assign to the variable 'coord.dist'
coord.dist <- dist(as.matrix(lats,lons))

# 4) Transform environmental data to numeric type, and assign to variable 'x1'
x1 <- as.numeric(Ponds$"SpC")

# 5) Using the `vegdist()` function in `simba`, calculate the Euclidean distance between the plots for environmental variables. Assign the result to the variable 'env.dist'
env.dist <- vegdist(x1, "euclidean") 
# 6) Transform all distance matrices into database format using the `liste()` function in `simba`:
comm.dist.ls <- liste(comm.dist, entry="comm") 
env.dist.ls <- liste(env.dist, entry="env") 
coord.dist.ls <- liste(coord.dist, entry="dist")

# 7) Create a data frame containing similarity of the environment and similarity of community.
df <- data.frame(coord.dist.ls, env.dist.ls[,3], comm.dist.ls[,3])
# 8) Attach the columns labels 'env' and 'struc' to the dataframe you just made.
names (df)[4:5] <- c("env","struc")
# 9) After setting the plot parameters, plot the distance-decay relationships, with regression lines in red.
par(mfrow=c(1, 2), pty="s")
plot(env, struc, xlab="Environmental Distance", ylab="1 - Bray-Curtis",
     main = "Environment", col='SteelBlue')
# 10) Use `simba` to calculates the difference in slope or intercept of two regression lines
OLS <- lm(struc ~ env)
OLS
abline(OLS, col = "red4")

plot(dist, struc = )

```


***Question 7***: What can you conclude about community similarity with regars to environmental distance and geographic distance?

> ***Answer 7***: 


## 7) EXAMINING SPECIES SPATIAL ABUNDANCE DISTRIBUTIONS

***Question 8***: In your own words, explain the species spatial abundance distribution and what it reveals.

> ***Answer 8***:


In the R code chunk below, do the following: 

1. Define a function that will generate the SSAD for a given OTU.
2. Draw six OTUs at random from the IN ponds dataset and and plot their SSADs as kernel density curves. Use **while loops** and **if** statements to accomplish this.

```{r}
# 1. Define an SSAD function

# 2. Set plot parameters

# 3. Declare a counter variable

# 4. Write a while loop to plot the SSADs of six species chosen at random 

```


## 8) UNDERSTANDING SPATIAL SCALE
Many patterns of biodiversity relate to spatial scale.

***Question 9***:  List, describe, and give examples of the two main aspects of spatial scale

> ***Answer 9***: 


## 9) CONSTRUCTING THE SPECIES-AREA RELATIONSHIP

***Question 10***: In your own words, describe the species-area relationship.

> ***Answer 10***: 


In the R code chunk below, provide the code to: 

1. Simulate the spatial distribution of a community with 100 species, letting each species have between 1 and 1,000 individuals.

```{r, results = 'hide', message = FALSE, warning = FALSE}
# 1. Declare variables to hold simulated community and species information 

# 2. Populate the simulated landscape


```

While consult the handout for assistance, in the R chunk below, provide the code to: 

1. Use a nested design to examine the SAR of our simulated community.
2. Plot the SAR and regression line. 

```{r}
# 1. Declare the spatial extent and lists to hold species richness and area data

# 2. Construct a 'while' loop and 'for' loop combination to quantify the numbers of species for progressively larger areas of the simulated landscape.

# 3. Be sure to log10-transform the richness and area data
```


In the R code chunk below, provide the code to: 

1. Plot the richness and area data as a scatter plot.
2. Calculate and plot the regression line
3. Add a legend for the z-value (i.e., slope of the SAR)

```{r, results = 'hide', message = FALSE, warning = FALSE}

```


***Question 10a***: Describe how richness relates to area in our simulated data by interpreting the slope of the SAR.

> ***Answer 10a***: 

***Question 10b***: What does the y-intercept of the SAR represent?

> ***Answer 10b***: 


## SYNTHESIS
Load the dataset you are using for your project. Plot and discuss either the geogrpahic Distance-Decay relationship, the SSADs for at least four species, or any variant of the SAR (e.g., random accumulation of plots or areas, accumulation of contiguous plots or areas, nested design).