---
title: 'Assignment: Spatial Diversity'
author: 'Ashwini Ramesh; Z620: Quantitative Biodiversity, Indiana University'
output:
  pdf_document: default
  html_document: default
geometry: margin=2.54cm
---

## OVERVIEW
This assignment will emphasize primary concepts and patterns associated with spatial diversity, while using R as a Geographic Information Systems (GIS) environment. 
Complete the assignment by refering to examples in the handout.

After completing this assignment you will be able to:  
1.  Begin using R as a geographical information systems (GIS) environment.  
2.  Identify primary concepts and patterns of spatial diversity.  
3.  Examine effects of geographic distance on community similarity.  
4.  Generate simulated spatial data.

## Directions:
1. Change "Student Name" on line 3 (above) with your name.
2. Complete as much of the assignment as possible during class; what you do not complete in class will need to be done on your own outside of class.
3. Use the handout as a guide; it contains a more complete description of data sets along with the proper scripting needed to carry out the assignment.
4. Be sure to **answer the questions** in this assignment document.
Space for your answer is provided in this document and indicated by the ">" character.
If you need a second paragraph be sure to start the first line with ">".
5. Before you leave the classroom, **push** this file to your GitHub repo.
6. When you are done wit the assignment, **Knit** the text and code into an html file.
7. After Knitting, please submit the completed assignment by creating a **pull request** via GitHub.
Your pull request should include this file *spatial_assignment.Rmd* and the html output of `Knitr` (*spatial_assignment.html*).


## 1) R SETUP
In the R code chunk below, provide the code to:  

1. Clear your R environment
2. Print your current working directory,
3. Set your working directory to your "*/Week4-Spatial*" folder, and

```{r, results = 'hide'}
rm(list=ls())
getwd() 
#setwd("~/GitHub/QuantitativeBiodiversity/QB-2017/Week4-Spatial")
```

## 2) LOADING R PACKAGES

In the R code chunk below, do the following:  

1. Install and/or load the following packages: `vegan`, `sp`, `gstat`, `raster`, `RgoogleMaps`, `maptools`, `rgdal`, `simba`, `gplots`, `rgeos`

```{r, results = 'hide', message = FALSE, warning = FALSE} 
require(vegan)
#install.packages('sp') # Classes and methods for handling spatial data 
require(sp)
#install.packages('ggmap')
require(ggmap)
#install.packages('gstat') # Methods for geostatistical analyses
require(gstat)
#install.packages('raster') # Methods to create a RasterLayer object
require(raster)
#install.packages('RgoogleMaps') # For querying the Google server for staticmaps
require(RgoogleMaps)
#install.packages('maptools') # Tools for manipulating and reading geospatialdata
require(maptools)
#install.packages('rgdal') # Geospatial Data Abstraction Library
require(rgdal)
#install.packages('simba') # Similarity measures for community data
require(simba)
#install.packages('gplots') # Programming tools for plotting data
require(gplots)
#install.packages('rgeos') # Geostatistical package, used here for semivariograms
require(rgeos)
require(rgdal)
```

***Question 1***:  What are the packages `simba`, `sp`, and `rgdal` used for?
  
> ***Answer 1***:  simba is used to measure similarity measures for community data. sp for the classes and methods for handling spatial data and rgdal for geosatial data absraction library


## 3) LOADING DATA
In the R code chunk below, use the example in the handout to do the following:  

1. Load the Site-by-Species matrix for the Indiana ponds datasets: BrownCoData/SiteBySpecies.csv
2. Load the Environmental data matrix: BrownCoData/20130801_PondDataMod.csv
3. Assign the operational taxonomic units (OTUs) to a variable 'otu.names'
4. Remove the first column (i.e., site names) from the OTU matrix.

```{r, results = 'hide', message = FALSE, warning = FALSE} 
#1. Load the Site-by-Species matrix for the Indiana ponds datasets: BrownCoData/SiteBySpecies.csv
Ponds <- read.table(file = "BrownCoData/20130801_PondDataMod.csv", head = TRUE, sep = ",") 

#2. Load the Environmental data matrix: BrownCoData/20130801_PondDataMod.csv
OTUs <- read.csv(file = "BrownCoData/SiteBySpecies.csv", head = TRUE, sep = ",")

#3. Assign the operational taxonomic units (OTUs) to a variable 'otu.names'
otu.names <- names(OTUs) # Get the names of the OTUs

#4. Remove the first column (i.e., site names) from the OTU matrix.
OTUs <- as.data.frame(OTUs[-1]) # remove first column (site names)

#Maximum species richness among each site
max(specnumber(OTUs))


```

***Question 2a***:  How many sites and OTUs are in the SiteBySpecies matrix? 

> ***Answer 2a***:  Sites = 51, OTUs = 16383

***Question 2b***:  What is the greatest species richness found among sites?

> ***Answer 2b***:  Site 1 : 3659


## 4) GENERATE MAPS

In the R code chunk below, do the following:  

1. Using the example in the handout, visualize the spatial distribution of our samples with a basic map in RStudio using the `GetMap` function in the package `RgoogleMaps`. 
This map will be centered on Brown County, Indiana (39.1 latitude, -86.3 longitude).

```{r}
lats <- as.numeric(Ponds[, 3]) # latitudes (north and south) 
lons <- as.numeric(Ponds[, 4]) # longitudes (east and west)
newmap <- GetMap(center = c(39.1,-86.3), zoom = 10, destfile = "PondsMap.png", maptype="terrain")
PlotOnStaticMap(newmap, zoom = 10, cex = 2, col = 'blue') # Plot map in RStudio 
PlotOnStaticMap(newmap, lats, lons, cex = 1, pch = 20, col = 'red', add = TRUE)
```

***Question 3***:  Briefly describe the geographical layout of our sites.

> ***Answer 3***:  - All points are roughly within Bloomington,IN
- Five aggregations/clumps found on the map
- These clumps are spread across fairly uneven surface
- Each clump has anywhere between 3 ~ 12 sampled points



In the R code chunk below, do the following:  

1. Using the example in the handout, build a map by combining lat-long data from our ponds with land cover data and data on the locations and shapes of surrounding water bodies.

```{r}
library(raster)
library(RgoogleMaps)
require(vegan)
#install.packages('sp') # Classes and methods for handling spatial data 
require(sp)
#install.packages('gstat') # Methods for geostatistical analyses
require(gstat)
#install.packages('raster') # Methods to create a RasterLayer object 
require(raster)
#install.packages('RgoogleMaps') # For querying the Google server for static maps. 
require(RgoogleMaps)
#install.packages('maptools') # Tools for manipulating and reading geospatial data 
require(maptools)
#install.packages('rgdal') # Geospatial Data Abstraction Library
require(rgdal)
#install.packages('simba') # Similarity measures for community data
require(simba)
#install.packages('gplots') # Programming tools for plotting data
require(gplots)
#install.packages('rgeos') # Geostatistical package, used here for semivariograms 
require(rgeos)
require(rgdal)

# 1. Import TreeCover.tif as a raster file.
Tree.Cover <- raster("TreeCover/TreeCover.tif") # import TreeCover.tif as a raster file.
# 2. Plot the % tree cover data
plot(Tree.Cover, xlab = 'Longitude', ylab = 'Latitude', main = 'Map of geospatial data for % tree cover, \nwaterbodies, and sample sites')
# 3. Import water bodies as a shapefile.
Water.Bodies <- readShapeSpatial("water/water.shp")
# 4. Plot the water bodies around our study area, i.e., Monroe County.
plot(Water.Bodies, border = 'cyan', axes = TRUE, add = TRUE)
# 5. Convert lat-long data for ponds to georeferenced points.
Refuge.Ponds <- SpatialPoints(cbind(lons,lats))
# 6. Plot the refuge pond locations
plot(Refuge.Ponds, line = 'r', col = "red", pch = 20, cex = 1.5, add = TRUE)
```

***Question 4a***: What are datums and projections?

> ***Answer 4a***:Datum: A feature of the different layers used in the map, such that it models/ mimic's the Earth shape in the same exact way.
Projections: The representation of a point from the spherical coordinate system on a 2-D surface


## 5) UNDERSTANDING SPATIAL AUTOCORRELATION

***Question 5***: In your own words, explain the concept of spatial autocorrelation.

> ***Answer 5***: It asks "How often/to what extent are spatial variables more likely to be found clustered in space?" If it is very likely, then it represents positive autocorrelation, and if spread out then it represents negative autocorrelation


## 6) EXAMINING DISTANCE-DECAY

***Question 6***: In your own words, explain what a distance decay pattern is and what it reveals.

> ***Answer 6***: Distance Decay: With increasing geographic distance, what is the rate of similarity between sites? It is touted that this is a decaying function over geographic distance. 
- What does it reveal? Communities/environments close to one another are more likely to be similar than ones that are further away from each other in space

In the R code chunk below, do the following:  

1. Generate the distance decay relationship for bacterial communities of our refuge ponds and for some of the environmental variables that were measured. 
Note: You will need to use some of the data transformations within the *semivariogram* section of the handout.

```{r, results = 'hide', message = FALSE, warning = FALSE}
##Read note attached at the end of this chunk

# Construct a new dataframe for coordinates
xy <- data.frame(pond.name = Ponds$Sample_ID, lats = Ponds$lat, lons = Ponds$long)
# Transform 'xy' into a spatial points dataframe
coordinates(xy) <- c("lats", "lons")
# Identify the current projection (i.e., lat-long) and datum (NAD83). In our case, the projection and datum
proj4string(xy) <- CRS("+proj=longlat + datum=NAD83")
# Transform the projection and data, so we can get meaningful distances. In this case, we will use the
UTM <- spTransform(xy, CRS("+proj=utm +zone=51 +ellps=WGS84"))
UTM <- as.data.frame(UTM)
Ponds$lat_utm <- UTM[,2] # lattitude data
Ponds$lon_utm <- UTM[,3] # longitude data
# coordinates(Ponds) = ~lats_utm+lons_utm
# vgm <- variogram(TDS~1, data=Ponds)
# vgm.fit = fit.variogram(vgm, model = vgm(1, "Sph", 900, 1,fit.sill=F, fit.range=F))
# plot(vgm, vgm.fit)

# 1) Calculate Bray-Curtis similarity between plots using the `vegdist()` function
comm.dist <- 1 - vegdist(OTUs) # Bray-Curtis similarity between the plots

# 2) Assign UTM lattitude and longitude data to 'lats' and 'lons' variables
lats <- as.numeric(Ponds$lats_utm) # lattitude data
lons <- as.numeric(Ponds$lons_utm) # longitude data

# 3) Calculate geographic distance between plots and assign to the variable 'coord.dist'
coord.dist <- dist(as.matrix(lats, lons)) # geographical distance between plots

# 4) Transform environmental data to numeric type, and assign to variable 'x1'
x1 <- as.numeric(Ponds$"SpC")

# 5) Using the `vegdist()` function in `simba`, calculate the Euclidean distance between the plots for environmental variables. Assign the result to the variable 'env.dist'
env.dist <- vegdist(x1, "euclidean")

# 6) Transform all distance matrices into database format using the `liste()` function in `simba`:
comm.dist.ls <- liste(comm.dist, entry="comm")
env.dist.ls <- liste(env.dist, entry="env")
coord.dist.ls <- liste(coord.dist, entry="dist")

# 7) Create a data frame containing similarity of the environment and similarity of community.
df <- data.frame(coord.dist.ls, env.dist.ls[,3], comm.dist.ls[,3])

# 8) Attach the columns labels 'env' and 'struc' to the dataframe you just made.
names(df)[4:5] <- c("env", "struc")
attach(df)

# 9) After setting the plot parameters, plot the distance-decay relationships, with regression lines in red.
par(mfrow=c(1, 2), pty="s")
plot(env, struc, xlab="Environmental Distance", ylab="1 - Bray-Curtis",
main = "Environment", col='SteelBlue')
OLS <- lm(struc ~ env)
OLS # print regression results to the screen
abline(OLS, col="red4")
plot(dist, struc, xlab="Geographic Distance", ylab="1 - Bray-Curtis",
main="Community\nComposition", col='darkorchid4')
OLS <- lm(struc ~ dist)
OLS # print regression results to the screen
abline(OLS, col="red4")

# 10) Use `simba` to calculates the difference in slope or intercept of two regression lines
diffslope(env, struc, dist, struc)


#Thi runs when I'm running the chunk individually on the script (Look at PNG attached in the uploaded folder.However, it refuses to knit on the HTML. Also, it does not give me the specific problematic line and just directs me to the first line of the chunk.  I'm commenting out the lines for the knit. 
```


***Question 7***: What can you conclude about community similarity with regards to environmental distance and geographic distance?

> ***Answer 7***: The community similarity experiences decay with increasing environmental distance when compared to the geographic distance. This means that communities that have more similar environmental conditions are more similar than those that differ. However, communities that are closer in the physical space are no different from communities that are far away from one another.

## 7) EXAMINING SPECIES SPATIAL ABUNDANCE DISTRIBUTIONS

***Question 8***: In your own words, explain the species spatial abundance distribution and what it reveals.

> ***Answer 8***:A SAD plot primarily asks: "At a given abundance, how many counts of a particular individual can one find?" Thus, SAD is a histogram of individual counts at varying abundances


In the R code chunk below, do the following: 

1. Define a function that will generate the SSAD for a given OTU.
2. Draw six OTUs at random from the IN ponds dataset and and plot their SSADs as kernel density curves. Use **while loops** and **if** statements to accomplish this.

```{r}
# 1. Define an SSAD function
ssad <- function(x){
ad <- c(2, 2)
ad <- OTUs[, otu]
ad = as.vector(t(x = ad))
ad = ad[ad > 0]
}

# 2. Set plot parameters
par(mfrow=c(2, 2))

# 3. Declare a counter variable
ct <- 0 # a counter variable
# 4. Write a while loop to plot the SSADs of six species chosen at random 
while (ct < 4){ # While the ct variable is less than 4, do ...
otu <- sample(1:length(OTUs), 1) # choose 1 random OTU
ad <- ssad(otu) # find the OTU's SSAD
if (length(ad) > 10 & sum(ad > 100)){ct <- ct + 1
plot(density(ad), col = 'red', xlab='Site abundance',
ylab='Probability Density', main = otu.names[otu])
}
}

```


## 8) UNDERSTANDING SPATIAL SCALE
Many patterns of biodiversity relate to spatial scale.

***Question 9***:  List, describe, and give examples of the two main aspects of spatial scale

> ***Answer 9***: Extent and Grain are the two main aspects of a spatial scale. While 'Extent' denotes the greatest distance accounted for in the study, 'Grain' denotes the smallest constituent unit of measurement in the study. 


## 9) CONSTRUCTING THE SPECIES-AREA RELATIONSHIP

***Question 10***: In your own words, describe the species-area relationship.

> ***Answer 10***: The species-area relationship states that with increasing sampling area one is more likely to find greater number of species associated with the area and that this pattern is dictated by the Arrhenius power law. 


In the R code chunk below, provide the code to: 

1. Simulate the spatial distribution of a community with 100 species, letting each species have between 1 and 1,000 individuals.

```{r, results = 'hide', message = FALSE, warning = FALSE}
# 1. Declare variables to hold simulated community and species information 
community <- c() # an initiall empty community
species <- c() # with zero species
# 2. Populate the simulated landscapeinitiate the plot, i.e., landscape
plot(0, 0, col='white', xlim = c(0, 100), ylim = c(0, 100),
xlab='x coordinate', ylab='y coordinate',
main='A simulated landscape occupied by 100
species, having 1 to 1000 individuals each.')
while (length(community) < 100){ # while the community has less than 100 species
# choose the mean, standard deviation, and species color at random
std <- runif(1, 1, 10)
ab <- sample(1000, 1)
x <- rnorm(ab, mean = runif(1, 0, 100), sd = std)
y <- rnorm(ab, mean = runif(1, 0, 100), sd = std)
color <- c(rgb(runif(1),runif(1),runif(1)))
points(x, y, pch=".", col=color)
species <- list(x, y, color)
community[[length(community)+1]] <- species
}

```

While consult the handout for assistance, in the R chunk below, provide the code to: 

1. Use a nested design to examine the SAR of our simulated community.
2. Plot the SAR and regression line. 

```{r}
# 1. Declare the spatial extent and lists to hold species richness and area data
lim <- 10 # smallest spatial extent. This also equals the spatial grain.
S.list <- c() # holds the number of species
A.list <- c() # holds the spatial scales

# 2. Construct a 'while' loop and 'for' loop combination to quantify the numbers of species for progressively larger areas of the simulated landscape.
while (lim <= 100){S <- 0 # initiate richness at zero
for (sp in community){ # for each species in the community
xs <- sp[[1]] # assign the x coordinates
ys <- sp[[2]] # assign the y coordinates
sp.name <- sp[[3]] # assign the species name
xy.coords <- cbind(xs, ys) # combine the columns for x and y coordinates
for (xy in xy.coords){ # for each pair of xy coordinates
if (max(xy) <= lim){ # if the individual is within our current spatial extent...
S <- S + 1 # then the species occurs there
break # break out of the last for loop because we now know the species occurs inside our samplign }
}
}# 3. Be sure to log10-transform the richness and area data 
S.list <- c(S.list, log10(S))
A.list <- c(A.list, log10(lim^2))
lim <- lim * 2 # increase the extent multiplicately
}}

```


In the R code chunk below, provide the code to: 

1. Plot the richness and area data as a scatter plot.
2. Calculate and plot the regression line
3. Add a legend for the z-value (i.e., slope of the SAR)

```{r, results = 'hide', message = FALSE, warning = FALSE}
results <- lm(S.list ~ A.list)
plot(A.list, S.list, col="dark red", pch=20, cex=2,
main="Species-area relationship",
xlab='ln(Area)', ylab='ln(Richness)')
abline(results, col="red", lwd=2)
int <- round(results[[1]][[1]],2)
z <- round(results[[1]][[2]],2)
legend(x=2, y=2, paste(c('slope = ', z), collapse = " "), cex=0.8,
box.lty=0)
```


***Question 10a***: Describe how richness relates to area in our simulated data by interpreting the slope of the SAR.

> ***Answer 10a***: 
The simulation reveals that the slope of the relationship is 0.02 i.e.((y2 - y1)/ (x2 - x1)) = 0.02. This means than for a difference of every 100 units on the x-axis(or sampling area) there is a corresponding increase of 2 units on the y-axis(or species richness scale). Here all scales are log-transformed. 

***Question 10b***: What does the y-intercept of the SAR represent?

> ***Answer 10b***: The y-intercept of SAR ~ 0.8 and the equation of the line is y = 0.02x + 0.8. Essentially, here the modelling is saying that for 0 area coverage the species richness is 0.8. Biologically this has no meaning. However, overall one can gather that, given that this relationship here is positive, the minimum species richness at any given area will always be 0.8.  


## SYNTHESIS
Load the dataset you are using for your project. Plot and discuss either the geogrpahic Distance-Decay relationship, the SSADs for at least four species, or any variant of the SAR (e.g., random accumulation of plots or areas, accumulation of contiguous plots or areas, nested design).

```{r, results = 'hide', message = FALSE, warning = FALSE}
SbyS_mammals <- read.table("SbyS_mammals.txt", quote = , sep = "\t", header = TRUE, fill = TRUE)
env_mammals <- read.table("MCDB_sites.csv", quote = "", sep = ",", header = TRUE, fill = TRUE)
xy_mammals <- data.frame(siteID = env_mammals$Site_ID, lats = env_mammals$Latitude, lons = env_mammals$Longitude)

# 1) Calculate Bray-Curtis similarity between plots using the `vegdist()` function
comm.dist_m <- 1 - vegdist(SbyS_mammals)
# 2) Assign UTM lattitude and longitude data to 'lats' and 'lons' variables
lats <- as.numeric(xy_mammals$lats)
lons <- as.numeric(xy_mammals$lons)
# 3) Calculate geographic distance between plots and assign to the variable 'coord.dist'
coord.dist_m <- dist(as.matrix(lats, lons))
# 6) Transform all distance matrices into database format using the `liste()` function in `simba`:
comm.dist.ls_m <- liste(comm.dist_m, entry = "comm")
coord.dist.ls_m <- liste(coord.dist_m, entry = "dist")
# 7) Create a data frame containing similarity of community.
df <- data.frame(coord.dist.ls_m, comm.dist.ls_m[,3])
# 8) Attach the columns labels 'struc' to the dataframe you just made.
names(df)[4] <- c("struc")
attach(df)
# 9) After setting the plot parameters, plot the distance-decay relationships, with regression lines in red.
par(mfrow = c(1,2), pty = "s")
plot(dist, struc, xlab = "Geographic Distance", ylab = "1 - Bray-Curtis", main = "Community\nComposition", col = "darkorchid4")
OLS <- lm(struc ~ dist)
OLS
abline(OLS, col = "red4")
```

> ***Answer SYNTHESIS***: The regression indicates that there is a negative relationship between geographic distance and BC for community composition. However, the R^2 values ~ 0.07 indicating this is a weak relationship and explains very little of the variation in data. Perhaps envrionmental conditions could better explain spatial auto-correlation in data. We are working on ways to extract this information from other sources.